Modeling mental simulation: a computational analysis of mental rotation

### 2014-01-29 man/cogsci-2014/cogsci-2014.tex
Work on intro/background, formatting references

\title{Modeling mental rotation}

@@

In this paper, we attempt to address these questions through a
rational analysis \cite{Marr:1983to,anderson90,Shepard:1987tt} of four
models of mental rotation. First, we discuss the previous literature
on mental imagery. Second, we outline computational- and
algorithmic-level analyses of the problem of mental rotation. Then, we
describe a behavioral experiment based on the classic mental rotation
studies \cite<e.g.,>{Cooper:1975wp}, and compare the results of the
experiment with each of the models. We conclude with a discussion of
the strengths and weaknesses of each model, and lay out directions for
future work.

@@

@@

The only model (of which the authors are aware) that has more
seriously attempted to address concerns regarding which simulation to
run is the parallel-process model of mental rotation by
\citeA{Funt:1983wn}. He determines the direction and axis of rotation
using the method-of-moments method; once this target rotation
computed, his model rotates one object until the target rotation is
reached, and then compares the two to see if they are
congruent. \TODO{?}

However, there is evidence that people do not actually rotate until
the objects are congruent, but rather pay attention to certain salient
features of the object, rather than a holistic representation
\cite{Just1976,Yuille:1982tx}. More recent evidence shows that when
people perform physical rotations, they do not rotate until congruency
is reached and may even rotate \textit{away} from near perfect matches
\cite{Gardony:2013gn}. Moreover, the state-of-the-art in computer
vision suggests that there is more to shape matching than simply
checking for congruency, particularly when shapes are complex or when
the shapes might not be exactly the same
\cite<e.g.,>{Belongie:2002tj,Sebastian:2003vm}.

so what are they actually doing?

perhaps they are performing mental manipulations of an image, but not
a single trajectory of rotation

@@

rational analysis \cite{Marr:1983to,anderson90,Shepard:1987tt}

1) the problem: determine something about the spatial transformations
that an object has undergone
2) the goal: make this determiniation while using the least amount of
resources
3) the solution: perform those rotations and reflections which will
give the most information (?) about the answer until an answer is
found

@@

we focus less on the representation, and more on the process, assuming
a spatial representation (because you need both to make a claim,
\cite{Anderson1978})

@@

eye tracking \cite{Just1976}

@@

Additionally, we assume that the observed shapes may only be
transformed by a small amount at a time, and each transformation is
costly in terms of computational resources. Specifically, if the
current mental image is $\Xt$, then:
\begin{equation}
  \mathbf{X}_{t+1} = \left\{ \begin{array}{ll}
      f(\Xt, 0, |\epsilon|) &\mbox{ rotate by $\epsilon$,} \\
      f(\Xt, 1, 0) &\mbox{ flip,} \\
      f(\Xa, 0, 0) &\mbox{ reset, or} \\
      f(\Xa, 1, 0) &\mbox{ reset and flip,} \\
    \end{array} \right.
  \label{eq:actions}
\end{equation}
where $\epsilon\sim \mathcal{N}(0, \sigma_\epsilon)$ and
$\sigma_\epsilon=0.7$ is the standard deviation of the step size in
radians.

@@

While the previous few models have all focused on \textit{searching}
for the global maximum, the goal is really to just approximate
Eq.~\ref{eq:lh-ratio} while still obeying the constraint of sequential
rotations. Instead of perfoming a search, we could try maintaining a
probability distribution over our \textit{estimate}, and then choose
actions which are expected to improve the estimate of
Eq.~\ref{eq:lh-ratio}. This strategy has several obvious
benefits. First, it does not make assumptions about the shape or scale
of the similarity function. Second, by choosing to sample places which
are informative, this method implicitly minimizes the amount of
rotation, but because it is constrained to sequential rotations,
should still exhibit the linear effect of the angle.

@@

\subsection{General analysis}

>>>>

\title{Modeling mental simulation: a computational analysis of mental rotation}

@@

In this paper, we explore these questions through rational analysis
\cite{Marr:1983to,anderson90,Shepard:1987tt} and compare four models
of mental rotation. First, we discuss the previous literature on
mental imagery. Second, we outline computational- and
algorithmic-level analyses of the problem of mental rotation. Then, we
describe a behavioral experiment based on the classic mental rotation
studies \cite<e.g.,>{Cooper:1975wp}, and compare the results of the
experiment with each of the models. We conclude with a discussion of
the strengths and weaknesses of each model, and lay out directions for
future work.

@@

\TODO{say something here}

\subsection{Mechanisms behind mental rotation}

@@

The only model (of which the authors are aware) that has more
seriously attempted to address concerns regarding which simulation to
run is the parallel-process model of mental rotation by
\citeA{Funt:1983wn}. He determines the direction and axis of rotation
using the method-of-moments method; once this target rotation
computed, his model rotates one object until the target rotation is
reached, and then compares the two to see if they are congruent. This
is a somewhat unsatisfying explanation, however: why would it be
necessary to perform the rotation if the answer is already known?

Perhaps people do not engage in a single trajectory of rotation that
ends with congruency. Some evidence suggests that people do not
perform a holistic rotation until the objects are congruent, but
rather pay attention to certain salient features of the object
\cite{Just1976,Yuille:1982tx}. A more recent study shows that when
performing \textit{physical} rotations, people do not rotate until
congruency is reached, and may even rotate \textit{away} from near
perfect matches \cite{Gardony:2013gn}. Moreover, the state-of-the-art
in computer vision suggests that there is more to shape matching than
simply checking for congruency, particularly when shapes are complex
or when the shapes might not be exactly the same
\cite<e.g.,>{Belongie:2002tj,Sebastian:2003vm}.

@@

It may be useful to take a step back and try to analyze mental
rotation from using rational analysis
\cite{Marr:1983to,anderson90,Shepard:1987tt}. At the computational
level, we could say that the \textit{problem} is to determine
something about the spatial transformations that an object has
undergone. At the algorithmic level, we are constrained by the notion
that mental images must be transformed in an analog manner (or at
least in a way that is approximately analog), and that mental images
are time-consuming and effortful to generate. Thus, the \textit{goal}
is to make this determiniation while using the least amount of
resources. Due to the sequential constratiant, the minimum amount of
computation will coincide with the minimum angle of rotation. So, the
rational \textit{solution} will be something that only rotates through
the minimum angle.

@@

@@

@@

Additionally, we assume that the observed shapes may only be
transformed by a small amount at a time, and each transformation is
costly in terms of computational resources. Specifically, if the
current mental image is $\Xt$, then:
\begin{equation}
  \mathbf{X}_{t+1} = \left\{ \begin{array}{ll}
      f(\Xt, 0, |\epsilon|) &\mbox{ rotate by $\epsilon$,} \\
      f(\Xt, 1, 0) &\mbox{ flip,} \\
      f(\Xa, 0, 0) &\mbox{ reset, or} \\
      f(\Xa, 1, 0) &\mbox{ reset and flip,} \\
    \end{array} \right.
  \label{eq:actions}
\end{equation}
where $\epsilon\sim \mathcal{N}(0, \sigma_\epsilon)$ and
$\sigma_\epsilon=0.6$ is the standard deviation of the step size in
radians.

@@

While the previous few models have all focused on \textit{searching}
for the global maximum, the goal is really to just approximate
Eq.~\ref{eq:lh-ratio} while still obeying the constraint of sequential
rotations. Instead of perfoming a search, we could try maintaining a
probability distribution over our \textit{estimate} of
Eq.~\ref{eq:lh-ratio}, and then choose actions which are expected to
improve the accuracy of the estimate. This strategy has several
obvious benefits. First, it does not make assumptions about the shape
or scale of the similarity function. Second, by choosing to sample
places which are informative, this method implicitly minimizes the
amount of rotation, but because it is constrained to sequential
rotations, should still exhibit the linear effect of the angle.

@@
